---
title: "MATH-23C PROJECT - Analysis of /r/ForeverAlone Dataset"
author: by Indri N. Himawan
output: rmarkdown::html_vignette
---

First, let's load up this data onto R and load up the right libraries.

```{r warning=FALSE, results='hide', message=FALSE}  
library("fitdistrplus")
library("stringr")
library("stats4")
library("e1071")
library("ggplot2")

ForeverAlone <- read.csv("lib/foreveralone.csv")
```

***

# Income and Suicide

I want to see how many people in this subreddit has/n't attempted suicide depending on their income. In other words, does income level affect likelihood of suicide?

## Income and Suicide: using stacked barplot

```{r fig.width=7, fig.height=5}
count <- table(ForeverAlone$attempt_suicide, ForeverAlone$income)
orderedCount <- count[,c(1,2,3,8,10,11,12,13,4,5,6,7,9)]
barplot(orderedCount, main = "Number of suicide attempts by income level",
        xlab = "Income Range",
        ylab = "Number of suicide attempts",
        legend.text = TRUE,
        las = 1,
        col = c("blue", "red"))
```

To no surprise, it looks like those whose income is the lower end has a higher rate ot suicide attempts.From this data, we can tell that just because a person has a higher income, does not mean that they will never attempt suicide. I can see a number in the `$150,000 to $174,999` salary range to have had attempted suicide. 

## Income and Suicide: Logical Regression
```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}

# helper function 
findAverage <- function(row){
  incomeRange <- row[5]
  cleanRange <- str_remove_all(incomeRange, "[$,A-z]")
  nums <- as.numeric(strsplit(cleanRange, "\\s+")[[1]])
  if (length(nums) == 1){
    nums[1]
  } else {
    sum(nums)/length(nums)
  }
}

ForeverAlone$incomeAverage <- apply(ForeverAlone, 1, findAverage)
attemptedSuicide <- (as.numeric(ForeverAlone$attempt_suicide=="Yes")); head(attemptedSuicide)

income <- ForeverAlone$incomeAverage
plot(income, attemptedSuicide, xlab = "Income", ylab="suicide probability", main = "logical regression of suicide probability as a function of income")
b <- cov(income, attemptedSuicide)/var(income)
a <- mean(attemptedSuicide) - b*mean(income);a
abline(a, b, col = "red")

MLL<- function(alpha, beta) {
  -sum( log( exp(alpha+beta*income)/(1+exp(alpha+beta*income)) )*attemptedSuicide +
          log(1/(1+exp(alpha+beta*income)))*(1-attemptedSuicide) )
}

results<-mle(MLL, start = list(alpha = 0, beta = 0)) #an initial guess is required
results@coef
curve( exp(results@coef[1]+results@coef[2]*x)/ (1+exp(results@coef[1]+results@coef[2]*x)),col = "blue", add=TRUE)

# test for all the income values to see if it is a good model
uniqueIncomes <- sort(unique(income))
#  [1]      0.0   5000.5  14999.5  24999.5  34999.5  44999.5  62499.5  87499.5 112499.5 137499.5
# [11] 162499.5 187499.0 200000.0

findMean <- function (incomeToEvaluate) {
  index <- which(income == incomeToEvaluate); head(index)
  mean(attemptedSuicide[index])   #20%
}

res <- lapply(uniqueIncomes, findMean)
cleanRes <- numeric(length(res))
for (idx in 1: length(res)) {
  cleanRes[idx] = res[[idx]]
}

cleanRes
 # [1] 0.18125000 0.26000000 0.13793103 0.06818182 0.15384615 0.12500000 0.28571429 0.00000000
 # [9] 0.00000000 0.50000000 0.00000000 0.00000000 0.00000000

```

Unfortunately from these numbers it doesnt look very promising for it to be statistically significant.. It fluctuates.

## Income and Suicide: Permutation Test
Even though it is not promising to be statistically significant, let's see what happens if we apply Permutation Test

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
# Permutation test: The mean number of salary for those who  attempted suicide is:
AttemptedIncomeAvg <- mean(ForeverAlone$incomeAverage[which(ForeverAlone$attempt_suicide == "Yes")]) # 18088.21
NotAttemptedIncomeAvg <- mean(ForeverAlone$incomeAverage[which(ForeverAlone$attempt_suicide == "No")]) # 20357.94
observed <- NotAttemptedIncomeAvg - AttemptedIncomeAvg; observed # 2269.724

# now let's use a permutation test
N <- 10000
diffs <- numeric(N)
for (i in 1:N){
  Attempted <- sample(ForeverAlone$attempt_suicide) 
  AttemptedIncomeAvg <- mean(ForeverAlone$incomeAverage[which(Attempted == 'Yes')])
  NotAttemptedIncomeAvg <- mean(ForeverAlone$incomeAverage[which(Attempted == 'No')])
  diffs[i] <- NotAttemptedIncomeAvg - AttemptedIncomeAvg
}

pvalue <- (sum(diffs >= observed)+1)/(N+1); pvalue # 0.2918708
```

If we say that statistically significant means they would need to have pvalue of <= 5%, this p-value is quite big (30%). Even though it seems like the histogram shows some good correlation between income and suicide attempt. In fact, the p-value suggests that there is not enough evidence to reject the null hypothesis income and suicide is not independent. Meaning suicide may/may not be affected by income.

## Salary and Suicide: Conclusion
Higher income can reduce suicide rate. But it does not eliminate the likelihood of attempting suicide. And the relationship is not statistically significant.

***

# Friends and Suicide
I expect people who have least friends have a correlation to attempting who attempted suicide.

## Friends and Suicide: finding correlation using histogram and probability density graph overlaid on a histogram

```{r fig.width=7, fig.height=5}
SuicideFriends <- subset(ForeverAlone, attempt_suicide == "Yes", friends)
hist(SuicideFriends$friends, main = "Number of suicide attempts by # of friends",
     xlab ="Number of friends",
     ylab = "Number of suicide attempts",
     col = "cyan", breaks="FD")
```

From here you can see that number of friends can exponentially reduce the likelyhood of suicide. In fact, let's try to overlay a exponential distribution. I use the `fitdistrplus` library to obtain a rate.

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
rate <- 0.2652106
hist(SuicideFriends$friends, main = "Age of those who attempted suicide w/ probability",
     xlab ="Age of suicide",
     col = "cyan",
     probability = TRUE, breaks="FD")
curve(dexp(x, rate = rate), add=TRUE, col="red")
```

It looks pretty fitting! We can say that the more friends you are EXPONENTIALLY less likely you are to attempt suicide. Let's try to test if this matches with a exponential distribution using distribution function

```{r fig.show='hide', results='hide'}
breaks <- hist(SuicideFriends$friends, breaks="FD")$breaks
observed <- hist(SuicideFriends$friends, breaks="FD")$count
observed[4] <- sum(observed[4:5])
observed[5] <- sum(observed[6:13])
observed <- observed[-(6:13)]
numOfObservations <- sum(observed)
expected <- numOfObservations*c(pexp(2, rate),
                                pexp(4, rate)-pexp(2, rate),
                                pexp(6, rate)-pexp(4, rate),
                                pexp(10, rate)-pexp(6, rate),
                                1-pexp(10, rate))
Chi2 <-sum((observed-expected)^2/expected); Chi2 # 5.465886

# Since we have our expected total equal to actual equal
# and imposed the rate parameter, the df is 5-1-1 = 3
Pvalue <- pchisq(Chi2,3,lower.tail = FALSE); Pvalue # 0.1406933
```

This means that there is 14% chance that this result happens as a exponential distribution. With this value, there is not enough evidence against the null hypothesis that our data did come from a exponential distribution.

## Friends and Suicide: finding correlation using permutation test (simulation method)
Let's see if through permutation test, whether or not number of friends has relation to whether or not to attempt suicide. Note that in order to come up with a proper normal distribution, because this data has a strong positive skew, I would need to trim the data. Find if there is a p-value that is <=5% if skewness is between -1 and 1. 

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
# function to trim dataset based on a percentage
trimFriendData <- function (data, percentage) {
  newData <- data

  upper_quantile <- quantile(newData$friends, 1-percentage)
  lower_quantile <- quantile(newData$friends, percentage)

  newData$friends[newData$friends > upper_quantile] <- NA
  newData$friends[newData$friends < lower_quantile] <- NA

  subset(newData, !is.na(friends))
}

findPValueWithTrimmedFriends <- function (percentage) {
  ForeverAlone.Trimmed <- trimFriendData(ForeverAlone, percentage)

  cat("This is the trimmed mean = ", mean(ForeverAlone.Trimmed$friends), " \n")
  cat("This is the skewness for your trimmed subset = ", skewness(ForeverAlone.Trimmed$friends), "\n")

  # Now let's redo the permutation test with this new table
  AttemptedFriendNumAvg <- mean(ForeverAlone.Trimmed$friends[which(ForeverAlone.Trimmed$attempt_suicide == "Yes")])
  NotAttemptedFriendNumAvg <- mean(ForeverAlone.Trimmed$friends[which(ForeverAlone.Trimmed$attempt_suicide == "No")])
  observed <- NotAttemptedFriendNumAvg - AttemptedFriendNumAvg; observed

  # let's see if through permutation test, whether or not number of friends has relation to whether or not to attempt suicide
  N <- 10000
  diffs <- numeric(N)
  for (i in 1:N){
    Attempted <- sample(ForeverAlone.Trimmed$attempt_suicide)   # permuted suicide attempt column
    AttemptedFriendNumAvg <- mean(ForeverAlone.Trimmed$friends[which(Attempted == 'Yes')])
    NotAttemptedFriendNumAvg <- mean(ForeverAlone.Trimmed$friends[which(Attempted == 'No')])
    diffs[i] <- NotAttemptedFriendNumAvg - AttemptedFriendNumAvg
  }

  hist(diffs, breaks = "FD", probability = TRUE) # BEAUTIFUL! It looks like a proper normal distribution :)
  abline(v = observed, col = "red")

  pvalue <- (sum(diffs >= observed)+1)/(N+1)
  cat("This is your p-value = ", pvalue, "\n")
}

# test out numbers, see which one is has low skewness and see if the p-value can still be statistically significant
# after a couple of tries with many numbers... found that `0.11` is the sweetspot to keep
# the skewness low but find a statistically significant subset.

findPValueWithTrimmedFriends(0.11)
```

Found that with trimming the data at 0.11, the trimmed mean =  `3.390168`, skewness = `0.9407331` and p-value = `0.04479552`. If we define statistical significance as p-value <= 0.05), do reject that having friends and suicide attempt are independent. And even after making data not have a high skewness level (between -1 and 1), they are actually statistically significant!

## Friends and Suicide: Logical Regression
```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
getLogicalRegressionValues <- function (percentage) {
  ForeverAlone.Trimmed <- trimFriendData(ForeverAlone, percentage)

  cat("This is the skewness for your trimmed subset = ", skewness(ForeverAlone.Trimmed$friends), "\n")

  # Let's try to see if using logistic regression can make it seem like it is statistical significant
  attemptedSuicide <- (as.numeric(ForeverAlone.Trimmed$attempt_suicide=="Yes")); head(attemptedSuicide)

  friends <- ForeverAlone.Trimmed$friends
  plot(friends, attemptedSuicide, xlab="Number of friends", ylab="Suicide Probability", main = "Logical Regression of suicide probability as a function of number of friends")
  b <- cov(friends, attemptedSuicide)/var(income) # obtain slope
  # Find the intercept
  a <- mean(attemptedSuicide) - b*mean(friends);a
  #We can add this regression line to the plot of the data
  abline(a, b, col = "red")

  MLL<- function(alpha, beta) {
    -sum( log( exp(alpha+beta*friends)/(1+exp(alpha+beta*friends)) )*attemptedSuicide +
            log(1/(1+exp(alpha+beta*friends)))*(1-attemptedSuicide) )
  }

  #R has a function that will maximize this function of alpha and beta
  results<-mle(MLL, start = list(alpha = 0, beta = 0)) #an initial guess is required
  results@coef
  curve( exp(results@coef[1]+results@coef[2]*x)/ (1+exp(results@coef[1]+results@coef[2]*x)),col = "blue", add=TRUE)
  # shows slight decline as number of friends grow

  # test for all the income values to see if it is a good model
  uniqueNumOfFriends <- sort(unique(friends))
  findMean <- function (numOfFriendsToEvaluate) {
    index <- which(friends == numOfFriendsToEvaluate); head(index)
    mean(attemptedSuicide[index])   #20%
  }

  res <- lapply(uniqueNumOfFriends, findMean)
  cleanRes <- numeric(length(res))
  for (idx in 1: length(res)) {
    cleanRes[idx] = res[[idx]]
  }
  cleanRes

  data.frame(number_of_friends = uniqueNumOfFriends, mean_attempted_suicide=cleanRes)
}

regressionWithThreshold <- getLogicalRegressionValues(0.11);
```
It's strange because this logical regression value fluctuates a lot in this subset. But apparently it is statistically significant? Surprise!?

## Friends and Suicide: Showing CLT
This graph of people having friends is not normal distribution. So let's use CLT so see if its mean is normal distribution.

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
ForeverAlone.Trimmed <- trimFriendData(ForeverAlone, 0.11) # use this percentage as found above
friendsNumber <- ForeverAlone.Trimmed$friends
hist(friendsNumber, breaks="FD", probability = TRUE, xlab = "number of friends", ylab="suicide probability", main="density of suicide") # looks nothing like a normaldistribution.

nFriends <- length(friendsNumber) # the number of tested
#To find the best fitting normal distribution, compute the mean and variance of the data
mu <- mean(friendsNumber); mu # 3.390168
sigma <- sd(friendsNumber); sigma # 3.393507
curve(dnorm(x, mu, sigma), add = TRUE, col = "red")
```

Now it shows the histogram of the data with an overlay of graph of normal density function
with the same mean and variance

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
dec <- qnorm(seq(0.0, 1, by = 0.1), mu, sigma); dec   #11 bins
Exp <- rep(nFriends/10,10); Exp     #expected friends per bin

binScores <- numeric(10)
for (i in 1:10)
  binScores[i] <- sum((friendsNumber >= dec[i]) & (friendsNumber <= dec[i+1]) )

binScores
# [1]   0 111  54  42  44   0  27  45  36  46

chisq <- sum((binScores - Exp)^2/Exp); chisq # 214.8272
#We estimated two parameters, which costs two degrees of freedom
pValue <- pchisq(chisq, df = 7, lower.tail = FALSE); pValue # 8.26185e-43
```

The very small p value which makes sense because this is nothing like a normal distribution.

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
# Now repeat, using average score for a sample of 25 friends
N <- 2000; xbars <- numeric(N)
for (i in 1:N) xbars[i] <- mean(sample(friendsNumber,25))
hist(xbars, breaks = "FD", probability = TRUE)
# looks like a normal distribution here

#For the best fit, compute the mean and variance of the data
mu25 <- mean(xbars); mu25;  # around 3.140168
sig25 <- sd(xbars); sig25; # around 0.5961972
curve(dnorm(x, mu25, sig25), add = TRUE, col = "red") # looks good!

# Make a vector of deciles
dec <- qnorm(seq(0.0, 1, by = 0.1), mu25, sig25); dec   #11 bins
Exp <- rep(N/10,10); Exp     #expected means per bin

#Count how many means are in each bin
binmeans <- numeric(10)
for (i in 1:10)  binmeans[i] <- sum((xbars >= dec[i]) & (xbars <= dec[i+1]) ); binmeans
#  [1] 198 188 206 229 190 199 204 212 170 204

#Test for uniformity using chi square.
chisq <- sum((binmeans - Exp)^2/Exp); chisq #  11.01
# We estimated two parameters, which costs two degrees of freedom
pValue <- pchisq(chisq, df = 7, lower.tail = FALSE); pValue # 0.1381834;
```
Using 0.05 significance level, Since p-value 0.1381834 is much larger than 0.05 significance level, we do not reject the null hypothesis that the normal distribution was a good model for the dataset. The calculated chi-square value could easily have arisen by chance. This result illustrates the central-limit theorem: Start with any distribution of finite variance and construct the mean of many samples. The sample means will have a distribution that is approximately normal.

In this case the sample means is even more normal than the actual data.

## Friends and Suicide: Conclusion
Higher number of friends can exponentially reduce chance of suicide. If you know somebody who is suicidal, try to help them feel less lonely. Be their friend! If you are suicidal, try to identify your friends and reach out to them (it's a hard thing to do, I know. But data does not lie.)

***

# Depression and Suicide
Is depression and suicide correlated?

## Depression and Suicide: using Contingency table to find correlation
```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
observed <- table(ForeverAlone$depressed, ForeverAlone$attempt_suicide); observed
  #      No Yes
  # No  147  10
  # Yes 237  75
```
It's no surprise to me that those who are depressed are more likely to attempt suicide. Now let's try to compare with an expected contingency table.

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
pDepressed <- mean(ForeverAlone$depressed == 'Yes')
pAttempted <- mean(ForeverAlone$attempt_suicide == 'Yes')
expected <- nrow(ForeverAlone) * outer(c(1 - pDepressed, pDepressed), c(1 - pAttempted, pAttempted)); expected
#          [,1]     [,2]
# [1,] 128.5458 28.45416
# [2,] 255.4542 56.54584
#
# looks pretty close from observed!

chisq <- sum((observed-expected)^2/expected); chisq # 21.97367
pValue <- 1 - pchisq(chisq, 1); pValue # 2.764167e-06
# Using the built in chisq method...
chisq.test(ForeverAlone$depressed, ForeverAlone$attempt_suicide, correct=FALSE)
#
# 	Pearson's Chi-squared test
#
# data:  ForeverAlone$depressed and ForeverAlone$attempt_suicide
# X-squared = 21.974, df = 1, p-value = 2.764e-06
#
# (Same result)
```

With such a small p-value = `2.764e-06`, we reject the null hypothesis that depression and suicide attempt are independent. meaning there may be a correlation between those who are depressed and those who have attempted suicide.

## Depression and Suicide: Conclusion
Depression and suicide is correlated. Decreasing depression decreases likelihood of suicide. Efforts to reduce suicide also means efforts to reduce depression.

***

# Age and Suicide

What is the age has the highest rate of suicide?

## Age and Suicide: using confidence interval
```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
SuicideAge <- subset(ForeverAlone, attempt_suicide == "Yes", age)$age
populationMean <- mean(SuicideAge) # 24.24706
populationSd <- sd(SuicideAge) # 6.307414

n <- 10
N <- 10^4

means <- numeric(N)
stdSamplesT <- numeric(N)
for (i in 1:N){
  values <- sample(SuicideAge, n)
  sampleMean <- mean(values)
  sampleSd <- sd(values)
  means[i] <- sampleMean
  stdSamplesT[i] <- (sampleMean - populationMean)/(sampleSd/sqrt(n))
}

hist(means, probability = TRUE)
curve(dnorm(x, populationMean, populationSd/sqrt(10)), add=TRUE, col="red")

hist(stdSamplesT, probability = TRUE, breaks = "FD")
curve(dt(x, n-1), add=TRUE, col="red")

#Let's see how we do with a t confidence interval
missedL <- 0; missedU <- 0
plot(x =c(15,40), y = c(1,100), type = "n", xlab = "age", ylab = "", main="confidence interval for age of those who attempted suicide") #blank plot
N <-1000
counter <- 0
for (i in 1:N) {
  x <- sample(SuicideAge, n)
  L <- mean(x) + qt(0.025, n-1) * sd(x)/sqrt(n) #usually less than the true mean
  U <- mean(x) + qt(0.975, n-1) * sd(x)/sqrt(n) #usually greater than the true mean
  if (populationMean <L) missedL <- missedL + 1 # +1 if we were wrong
  if (populationMean >U) missedU <- missedU + 1 # +1 if we were wrong
  if (L < populationMean && U > populationMean) counter <- counter + 1 #count +1 if we were correct
  if(i <= 100) segments(L, i, U, i)
}
abline (v = populationMean, col = "red") #vertical line at true mean
(N-missedL-missedU)/N # 0.93
missedL/N # 0.006
missedU/N #0.064
```
The fraction of the time did interval include the true mean is `0.93`. Which is significant. However, not all intervals include the true mean. The confidence interval rarely lies to the left of the true mean `0.006` but a bit more often on the right `0.064`.

## Age and Suicide: conclusion
The confidence interval shows that `0.93` of all intervals include the true mean. This is good because if there are campaigns for suicide prevention, they can be directed to those in their mid-twenties.

***

# Age, Sexuality and Suicide

Do those who are part of the LGBTQ+ community struggle more with suicide? 

## Age, Sexuality, and Suicide: using ggplot
```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
ForeverAlone.SuicideSubset <- subset(ForeverAlone, attempt_suicide == "Yes")
ForeverAlone.SuicideSubset$isStraight <- ifelse(ForeverAlone.SuicideSubset$sexuallity == 'Straight', 1, 0)
ForeverAlone.SuicideSubset$isStraight <- factor(ForeverAlone.SuicideSubset$isStraight,levels=c(0,1),
   labels=c("Gay/Lesbian/Bisexual","Straight"))
qplot(age, data=ForeverAlone.SuicideSubset, geom="density", fill=isStraight, alpha=I(.5),
   main="Distribution of Suicide rate by sexuality", xlab="Age",
   ylab="Density")
```

Here it is obvious that those do not identify as straight struggle more with suicide. 

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
qplot(isStraight, age, data=ForeverAlone.SuicideSubset, geom=c("boxplot", "jitter"),
   fill=isStraight, main="Age of those attempted suicide by sexuality",
   xlab="", ylab="Age")
```

Both are centered more in mid-20s but for those who do not identify as straight, the distribution is even more condensed in their mid-20s.

```{r fig.width=7, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
ForeverAlone$hasSuicide <- ifelse(ForeverAlone$attempt_suicide == "Yes", 1, 0)
ForeverAlone$isStraight <- ifelse(ForeverAlone$sexuallity == 'Straight', 1, 0)
ForeverAlone$isStraight <- factor(ForeverAlone$isStraight,levels=c(0,1),
   labels=c("Gay/Lesbian/Bisexual","Straight"))

qplot(age, hasSuicide, data=ForeverAlone, geom=c("point", "smooth"),
   method="lm", formula=y~x, color=isStraight,
   main="Regression of Suicide density on Age divided by sexuality",
   xlab="Age", ylab="Suicide density")
```

From all the results, I am seeing an overwhelming amount of people who are part of the LGBTQ community are more likely to attempt suicide.

## Age, Sexuality and Suicide : conclusion
Finding that the hotspot of people who attempted suicide is in their mid-20s and that those who are part of the LGBT has higher risk of suicide.

***

# Overall conclusion
Having more income does not fix all your problems if you've already been struggling with suicide. Of course, having basic necessities would help. But being rich does not make you completely happy. However, having friends can exponentially lessen likelyhood of suicide. There is also a strong correlation between suicide probability and depression. Those who experience depression and more likely to to attempt suicide. Treating  depression can be instrumental in reducing suicide attempts.

Aside from this, people who attempt suicide are more likely in their 20s. Those who do not identify as straight have a higher risk of suicide.

End of document.