---
title: "MATH-23C PROJECT - Analysis of /r/ForeverAlone Dataset"
author: by Indri N. Himawan
output: rmarkdown::html_vignette
---

First, let's load up this data onto R.

```{r}
ForeverAlone <- read.csv("lib/foreveralone.csv")
```

***

# Salary and Suicide

I want to see how many people in this subreddit has/n't attempted suicide depending on their income. In other words, does income level affect likelihood of suicide?

## Salary and Suicide: finding correlation using a barplot.

```{r fig.width=8, fig.height=5}
count <- table(ForeverAlone$attempt_suicide, ForeverAlone$income)
orderedCount <- count[,c(1,2,3,8,10,11,12,13,4,5,6,7,9)]
barplot(orderedCount, main = "Number of suicide attempts by income level",
        xlab = "Income Range",
        ylab = "Number of suicide attempts",
        legend.text = TRUE,
        las = 1,
        col = c("blue", "red"))
```

To no surprise, it looks like those whose income is the lower end has a higher rate ot suicide attempts. However, this could have also have come to light only because people who use the subreddit are mostly ones with lower income. From this data, we can tell that just because a person has a higher income, does not mean that they will never attempt suicide. I can see a number in the `$150,000 to $174,999` salary range to have had attempted suicide. 

## Salary and Suicide: Conclusion
Higher income can reduce suicide rate. But it does not eliminate the likelihood of attempting suicide.

***

# Friends and Suicide
Another obvious thing that can influence a person's decision to die by suicide is the number of friends. The more alone a person is, the more they feel like it won't matter if they passed. I expect people who have least friends have a correlation to attempting who attempted suicide.

## Friends and Suicide: finding correlation using histogram and probability density graph overlaid on a histogram

```{r fig.width=8, fig.height=5}
SuicideFriends <- subset(ForeverAlone, attempt_suicide == "Yes", friends)
hist(SuicideFriends$friends, main = "Number of suicide attempts by # of friends",
     xlab ="Number of friends",
     ylab = "Number of suicide attempts",
     col = "cyan", breaks="FD")
```

From here you can see that number of friends can exponentially reduce the likelyhood of suicide. In fact, let's try to overlay a exponential distribution. I use the `fitdistrplus` library to obtain a rate.

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
rate <- 0.2652106
hist(SuicideFriends$friends, main = "Age of those who attempted suicide w/ probability",
     xlab ="Age of suicide",
     col = "cyan",
     probability = TRUE, breaks="FD")
curve(dexp(x, rate = rate), add=TRUE, col="red")
```

It looks pretty fitting! We can say that the more friends you are EXPONENTIALLY less likely you are to attempt suicide. Let's try to test if this matches with a exponential distribution using distribution function

```{r fig.show='hide', results='hide'}
breaks <- hist(SuicideFriends$friends, breaks="FD")$breaks
observed <- hist(SuicideFriends$friends, breaks="FD")$count
observed[4] <- sum(observed[4:5])
observed[5] <- sum(observed[6:13])
observed <- observed[-(6:13)]
numOfObservations <- sum(observed)
expected <- numOfObservations*c(pexp(2, rate),
                                pexp(4, rate)-pexp(2, rate),
                                pexp(6, rate)-pexp(4, rate),
                                pexp(10, rate)-pexp(6, rate),
                                1-pexp(10, rate))
Chi2 <-sum((observed-expected)^2/expected); Chi2 # 5.465886

# Since we have our expected total equal to actual equal
# and imposed the rate parameter, the df is 5-1-1 = 3
Pvalue <- pchisq(Chi2,3,lower.tail = FALSE); Pvalue # 0.1406933
```

This means that there is 14% chance that this result happens as a exponential distribution. With this value, there is not enough evidence against the null hypothesis that our data did come from a exponential distribution.

## Friends and Suicide: finding correlation using permutation test (simulation method)
Let's see if through permutation test, whether or not number of friends has relation to whether or not to attempt suicide. Note that in order to come up with a proper normal distribution, I would need to get rid of outliers. Because the data was giving a max of `600` friends but a median of `3`.

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
ForeverAlone.NoFriendOutliers <- subset(ForeverAlone, friends <= 10) # getting rid of outliers
AttemptedFriendNumAvg <- mean(ForeverAlone.NoFriendOutliers$friends[
        which(ForeverAlone.NoFriendOutliers$attempt_suicide == "Yes")])
NotAttemptedFriendNumAvg <- mean(ForeverAlone.NoFriendOutliers$friends[
        which(ForeverAlone.NoFriendOutliers$attempt_suicide == "No")])
observed <- NotAttemptedFriendNumAvg - AttemptedFriendNumAvg

# using this simulation method
N <- 10000
diffs <- numeric(N)
for (i in 1:N){
  Attempted <- sample(ForeverAlone.NoFriendOutliers$attempt_suicide) 
  AttemptedFriendNumAvg <- mean(ForeverAlone.NoFriendOutliers$friends[
          which(Attempted == 'Yes')])
  NotAttemptedFriendNumAvg <- mean(ForeverAlone.NoFriendOutliers$friends[
          which(Attempted == 'No')])
  diffs[i] <- NotAttemptedFriendNumAvg - AttemptedFriendNumAvg
}

hist(diffs, breaks = "FD", probability = TRUE)
abline(v = observed, col = "red")

pvalue <- (sum(diffs >= observed)+1)/(N+1); pvalue # around 0.08559144
```

If we're using `0.1` significations level, do reject that having friends and suicide attempt are independent. Meaning that there is some correlation, which is no surprise to me. In fact, if we do not get rid of the outliers, the p-value would result to `0.00419958`, which enforces the correlation.

## Friends and Suicide: Using CLT
This graph of people having friends is not normal distribution. So let's use CLT so see if its mean is normal distribution.

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
friendsNumber <- ForeverAlone.NoFriendOutliers$friends
hist(friendsNumber, breaks="FD", probability = TRUE) 
nFriends <- length(friendsNumber) # the number of tested
mu <- mean(friendsNumber); mu # 3.130123
sigma <- sd(friendsNumber); sigma # 3.081096
curve(dnorm(x, mu, sigma), add = TRUE, col = "red")
```

Now it shows the histogram of the data with an overlay of graph of normal density function
with the same mean and variance

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
dec <- qnorm(seq(0.0, 1, by = 0.1), mu, sigma); dec   #11 bins
Exp <- rep(nFriends/10,10); Exp     #expected friends per bin

binScores <- numeric(10)
for (i in 1:10)
  binScores[i] <- sum((friendsNumber >= dec[i]) & (friendsNumber <= dec[i+1]) )

binScores
# [1]   0 111  54  42  44   0  27  45  36  46

chisq <- sum((binScores - Exp)^2/Exp); chisq # 214.8272
#We estimated two parameters, which costs two degrees of freedom
pValue <- pchisq(chisq, df = 7, lower.tail = FALSE); pValue # 8.26185e-43
```

The very small p value which makes sense because this is nothing like a normal distribution.

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
# Now repeat, using average score for a sample of 25 friends
N <- 2000; xbars <- numeric(N)
for (i in 1:N) xbars[i] <- mean(sample(friendsNumber,25))
hist(xbars, breaks = "FD", probability = TRUE)
# looks like a normal distribution here

#For the best fit, compute the mean and variance of the data
mu25 <- mean(xbars); mu25;  # around 3.140168
sig25 <- sd(xbars); sig25; # around 0.5961972
curve(dnorm(x, mu25, sig25), add = TRUE, col = "red") # looks good!

# Make a vector of deciles
dec <- qnorm(seq(0.0, 1, by = 0.1), mu25, sig25); dec   #11 bins
Exp <- rep(N/10,10); Exp     #expected means per bin

#Count how many means are in each bin
binmeans <- numeric(10)
for (i in 1:10)  binmeans[i] <- sum((xbars >= dec[i]) & (xbars <= dec[i+1]) ); binmeans
#  [1] 198 188 206 229 190 199 204 212 170 204

#Test for uniformity using chi square.
chisq <- sum((binmeans - Exp)^2/Exp); chisq #  11.01
# We estimated two parameters, which costs two degrees of freedom
pValue <- pchisq(chisq, df = 7, lower.tail = FALSE); pValue # 0.1381834;
```
Using 0.05 significance level, Since p-value 0.1381834 is much larger than 0.05 significance level, we do not reject the null hypothesis that the normal distribution was a good model for the dataset. This time the calculated chi-square value could easily have arisen by chance
This result illustrates the central-limit theorem: Start with any distribution of finite variance and construct the mean of many samples. The sample means will have a distribution that is approximately normal.

In this case the sample means is even more normal than the actual data.

## Friends and Suicide: Conclusion
Higher number of friends can exponentially reduce chance of suicide. If you know somebody who is suicidal, try to help them feel less lonely. Be their friend! If you are suicidal, try to identify your friends and reach out to them (it's a hard thing to do, I know. But data does not lie.)

***

# Depression and Suicide
Is depression and suicide correlated?

## Depression and Suicide: using Contingency table to find correlation
```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
observed <- table(ForeverAlone$depressed, ForeverAlone$attempt_suicide); observed
  #      No Yes
  # No  147  10
  # Yes 237  75
```
It's no surprise to me that those who are depressed are more likely to attempt suicide. Now let's try to compare with an expected contingency table.

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
pDepressed <- mean(ForeverAlone$depressed == 'Yes')
pAttempted <- mean(ForeverAlone$attempt_suicide == 'Yes')
expected <- nrow(ForeverAlone) * outer(c(1 - pDepressed, pDepressed), c(1 - pAttempted, pAttempted)); expected
#          [,1]     [,2]
# [1,] 128.5458 28.45416
# [2,] 255.4542 56.54584
#
# looks pretty close from observed!

chisq <- sum((observed-expected)^2/expected); chisq # 21.97367
pValue <- 1 - pchisq(chisq, 1); pValue # 2.764167e-06
```

With such a small p-value = 2.764e-06, we reject the null hypothesis that depression and suicide attempt are independent. meaning there may be a correlation between those who are depressed and those who have attempted suicide.

```{r fig.width=8, fig.height=5, message=FALSE, warning=FALSE, results='hide'}
# Using the built in chisq method...
chisq.test(ForeverAlone$depressed, ForeverAlone$attempt_suicide, correct=FALSE)
#
# 	Pearson's Chi-squared test
#
# data:  ForeverAlone$depressed and ForeverAlone$attempt_suicide
# X-squared = 21.974, df = 1, p-value = 2.764e-06
#
# (Same result)
```
With such a small `p-value = 2.764e-06`, we reject the null hypothesis that depression and suicide attempt are independent. Meaning there may be a correlation between those who are depressed and those who have attempted suicide.

## Depression and Suicide: Conclusion
Depression and suicide is correlated.


